Light inextensible strings (thread) under tension in the Schwarzschild geometry

### 2021-02-27 tpe_edreview_rebuttal.txt

One can "go the other way"---that is, given ODE $r''=r+2(r')^2/r$, one
can deriving $r=r_0\sec\phi$---by substituting $u=1/r$ and obtaining
$u''+u=0$, whence $u=A\cos(\phi-\phi_0)$.

I have rephrased slightly and numbered one of the equations so I can
refer to it.

> 13.  Can the author explain more fully what is meant by a "tangential
> null geodesic" (several lines after Eq. 6)?  That terminology makes it
> sound like the r-coordinate does not change, whence $r'$ would be
> zero, but that appears to not be the case.  In this section, if $r'$
> is small I see where the approximation $r''=r-2M$ comes from, but I am
> having trouble seeing where $r''-r-3M$ arises for the alleged
> tangential null geodesics.  What am I not seeing?

Everything can be resolved and clarified by omitting the word
"tangential" which was misleading, and replacing "small" with
"negligible".  The sentence now reads "If $r'$ is negligible, then a
string follows $r''=r-2M$, while null geodesics follow $r''=r-3M$".

The $r''=r-3M$ is just the (new) equation 7 -viz
$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$ but with $r'=0$.

### 2021-02-28 tpe_edreview_rebuttal.txt

Herewith my rebuttal to the review of "Light inextensible strings..."
The referee's comments are preceded by a chevron (">") and my response
is inline.  I have tinkered with linebreaks here and there in the
interests of clarity, and have replaced some mathematical notation in
the original review with latex equivalents.

Short story: I agree with all the comments made and have modified the
manuscript accordingly.  The document is stronger, more cohesive, more
accurate, and hopefully more instructive.  I submit that the extra
clarity and pedagogical value more than compensates for the slight
extra length.

I have changed the discussion slightly (emphasising the
staticness of the metric considered) as a result of reflecting on the
totality of the comments too.

@@

> 10. Suggestion.  Line below Eq. (5), I think the word "dashes" should
> be "primes".  This is a dash: --- and this is a prime: ', at least in
> usage with which I am familiar.

Fixed, notation now consistent

> 
> 11.  The paragraph below Eq. (6) is puzzling.  Here the path of the
> tight string is being compared to an "equatorial null geodesic," which
> I take to mean the path of a light beam moving along the equator.  If
> we are considering a light beam, then shouldn't a term with $t$ in it
> appear somewhere?  The puzzlement continues given the statement that
> "light, being sensitive to the time component of the metric" whereas
> the equations discussed in this paragraph say nothing (that I can see)
> about the time component of the metric.  The proper time interval
> would be zero, but the coordinate time interval $dt$ would, I think,
> be nonzero.  What am I missing?
>

The equation given as $u''=3u^2/2-u$ gives the path of a beam of light
through Schwarzschild space in which $r$ is given as a function of
angle, $r=r(\phi)$.  The idea is that a stationary observer has a
laser pointer pointing in a particular direction, switches it on and
leaves it on for a long time.  We then record all the points that the
beam touches.  The path of the light is then simply $r=r(\phi)$, no
time coordinate involved, as Figures 4 and 6 show.  I have changed the
citation to Rindler, who gives a detailed derivation.

@@

> 13.  Can the author explain more fully what is meant by a "tangential
> null geodesic" (several lines after Eq. 6)?  That terminology makes it
> sound like the r-coordinate does not change, whence $r'$ would be
> zero, but that appears to not be the case.  In this section, if $r'$
> is small I see where the approximation $r''=r-2M$ comes from, but I am
> having trouble seeing where $r''-r-3M$ arises for the alleged
> tangential null geodesics.  What am I not seeing?

Everything can be resolved and clarified by omitting the word
"tangential" which was misleading, and replacing "small" with
"negligible".  The sentence now reads "If $r'$ is negligible, then a
string follows $r''=r-2M$, while null geodesics follow $r''=r-3M$".

The $r''=r-3M$ is just the (new) equation 7 -viz
$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$ but with $r'=0$.

> 
> I will have to take the author's word for the numerical results.
>

Not any more! All the software used to create the diagrams is now
publically available under version control at

https://github.com/RobinHankin/string

and I have added a citation to this repo.  The code is heavily
documented.

> This is an interesting paper.  I congratulate the author for thinking
> of such an intriguing application of the Schwarzschild metric.
> However, in this reviewer's opinion, before the paper can be well
> fitted to its announced audience ("an undergraduate homework problem")
> some points need to be clarified.

### 2021-02-28 string_physics_educator_revision3.tex
fixes #46
fixes #47
further minor rewording

\begin{abstract}
Light inextensible string under tension is a stalwart feature of
elementary physics.  Here I show how considering such string in the
vicinity of a black hole, with the help of computer algebra systems,
can generate insight into the Schwarzschild geometry in the context of
an undergraduate homework problem.  Light taut strings minimize their
proper length, given by integrating the spatial component of the
Schwarzschild metric along the string.  The path itself is given by
straightforward numerical solution to the Euler-Lagrange equations.
If the string is entirely outside the event horizon, its closest
approach to the singularity is tangential.  At this point the string
is {\em visibly} curved, surely a memorable and informative insight.
The geometry of the Schwarzschild metric induces some interesting
nonlocal phenomena: if the distance of closest approach is less than
about $1.0761\cdot 2M$, the string self-intersects, even though it is
everywhere under tension.  Light taut strings furnish a third
interpretation of the concept ``straight line'', the other two being
null geodesics and free-fall world lines.
\end{abstract}

@@

The simplest configuration for a string would be circular at
constant~$r$ whose proper length would be given by integration of
equation~\ref{schwarzschild} as~$\int_{\phi=0}^{2\pi}r\,d\phi=2\pi r$
(this is why $r$ is sometimes called ``circumferential radius'').
Note carefully that $r$ is not the distance {\em to} anything, so this
result---which operates in Schwarzschild geometry and not plane
Euclidean geometry---is less obvious than it might appear.
Considering next a purely radial string we may follow Taylor and
Wheeler \cite{taylor2000} and calculate the proper length
$\ell\left(r_1,r_2\right)$ of a string stretching from~$r=r_1\geqslant
2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:

@@

There does not appear to be a simple analytical solution to
equation~\ref{rprimeprime}; but it is a straightforward second order
ODE, amenable to numerical solution and no harder than the equivalent
for null geodesics, a standard homework problem.  Here we consider
only strings exterior to the event horizon, $r>2M$, which have a
well-defined point of closest approach to the singularity.  Because
the system is invariant under~$\phi\longrightarrow-\phi$, the string
must be symmetrical about this point.  If $r'$ is small, then a string
follows $r''=r-2M$, while tangential null geodesics follow $r''=r-3M$.
Working to first order, and writing~$d$ for the difference in
Schwarzschild~$r$ between string and null geodesic, we have $d''=M$.
Transforming coordinates to $(x,y) = (r\phi,r-r_0)$ and with boundary
conditions corresponding to both string and null geodesic being
tangential at the $(x,y)$ origin, we see that
$d=\frac{M}{2}(x/r_0)^2$.  However, a static local observer tied to
the string would measure
$d_\mathrm{local}=d/\sqrt{1-2M/r_0}$---applying
equation~\ref{schwarzschild}---and obtain
$d_\mathrm{local}=\frac{M}{2}(x/r_0)^2\left(1-2M/r_0\right)^{-1/2}$
which would correspond to a radius of curvature
of~$r_0\cdot\frac{r_0}{M}\cdot\left(1-2M/r_0\right)^{-1/2}$.  Thus a
taut string would {\em look} curved to a static observer in
Schwarzschild coordinates, in the sense that a tangential light beam
and taut string would diverge quadratically.  At the photon sphere
$r_0=3M$, such a string would appear to have a radius of~$\sqrt{27}M$.
For a solar mass black hole this would be about $15\unit{km}$, a
readily understandable result.

@@

Here I solve equation~\ref{rprimeprime} numerically, using the R
programming language~\cite{rcore2019} and the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  First we
consider strings for which $r>2M$ everywhere (that is, completely
exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rprimeprime}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.

@@

\begin{figurehere} % Created by "light_closest_approach.R"
  \centerline{
    \includegraphics[width=\linewidth]{light_closest_approach.pdf}
  }
  \caption{Null geodesics in the Schwarzschild geometry, tangential
    at~$\phi=0$.  Note the differences between these curves and the taut
    strings shown elsewhere: unlike taut strings, null geodesices may
    cross the event horizon inwards, and are never tangential to the
    event horizon}
  \label{light_closest_approach}
\end{figurehere}

@@

The concept of ``straight line'' is problematic in general relativity.
Although photon paths provide a definition of straightness, their
world lines are sensitive to the time component of the metric.  Light
taut strings respond only to the spatial components of the metric, and
their configuration is arguably a more intuitive and helpful way to
think about straightness.  Using numerical methods to compute string
configuration is suitable for an undergraduate homework problem.

>>>>

\begin{abstract}
Light inextensible string under tension is a stalwart feature of
elementary physics.  Here I show how considering such string in the
vicinity of a black hole, with the help of computer algebra systems,
can generate insight into the Schwarzschild geometry in the context of
an undergraduate homework problem.  Light taut strings minimize their
proper length, given by integrating the spatial component of the
Schwarzschild metric along the string.  The path itself is given by
straightforward numerical solution to the Euler-Lagrange equations.
If the string is entirely outside the event horizon, its closest
approach to the singularity is tangential.  At this point the string
is {\em visibly} curved, surely a memorable and informative insight.
The geometry of the Schwarzschild metric induces some interesting
nonlocal phenomena: if the distance of closest approach is less than
about $1.0761\cdot 2M$, the string self-intersects, even though it is
everywhere under tension.  Light taut strings furnish a third
interpretation of the concept ``straight line'', the other two being
null geodesics and free-fall world lines.  All the software used is
available under the GPL~\cite{hankin2021}.
\end{abstract}

@@

The simplest configuration for a string would be circular at
constant~$r$ whose proper length is given by integration of
equation~\ref{schwarzschild} as~$\int_{\phi=0}^{2\pi}r\,d\phi=2\pi r$
(this is why $r$ is sometimes called ``circumferential radius'').
Note carefully that $r$ is not the distance {\em to} anything, so this
result---which operates in Schwarzschild geometry and not plane
Euclidean geometry---is less obvious than it might appear.
Considering next a purely radial string we may follow Taylor and
Wheeler \cite{taylor2000} and calculate the proper length
$\ell\left(r_1,r_2\right)$ of a string stretching from~$r=r_1\geqslant
2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:

@@

There does not appear to be a simple analytical solution to
equation~\ref{rprimeprime}; but it is a straightforward second order
ODE, amenable to numerical solution and no harder than the equivalent
for null geodesics, a standard homework problem.  Here we consider
only strings exterior to the event horizon, $r>2M$, which have a
well-defined point of closest approach to the singularity.  Because
the system is invariant under~$\phi\longrightarrow-\phi$, the string
must be symmetrical about this point.  If $r'$ is negligible, then a string
follows $r''=r-2M$, while null geodesics follow $r''=r-3M$.
Working to first order, and writing~$d$ for the difference in
Schwarzschild~$r$ between string and null geodesic, we have $d''=M$.
Transforming coordinates to $(x,y) = (r\phi,r-r_0)$ and with boundary
conditions corresponding to both string and null geodesic being
tangential at the $(x,y)$ origin, we see that
$d=\frac{M}{2}(x/r_0)^2$.  However, a static local observer tied to
the string would measure
$d_\mathrm{local}=d/\sqrt{1-2M/r_0}$---applying
equation~\ref{schwarzschild}---and obtain
$d_\mathrm{local}=\frac{M}{2}(x/r_0)^2\left(1-2M/r_0\right)^{-1/2}$
which would correspond to a radius of curvature
of~$r_0\cdot\frac{r_0}{M}\cdot\left(1-2M/r_0\right)^{-1/2}$.  Thus a
taut string would {\em look} curved to a static observer in
Schwarzschild coordinates, in the sense that a tangential light beam
and taut string would diverge quadratically.  At the photon sphere
$r_0=3M$, such a string would appear to have a radius of~$\sqrt{27}M$.
For a solar mass black hole this would be about $15\unit{km}$, a
readily understandable result.

@@

Here I solve equation~\ref{rprimeprime} numerically, using the R
programming language~\cite{rcore2019} and the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  We consider
only strings for which $r>2M$ everywhere (that is, completely exterior
strings).  Figure~\ref{closest_approach_non_self_intersecting} shows a
sequence of non-self-intersecting numerical solutions to
equation~\ref{rprimeprime}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.

@@

\begin{figurehere} % Created by "light_closest_approach.R"
  \centerline{
    \includegraphics[width=\linewidth]{light_closest_approach.pdf}
  }
  \caption{Null geodesics in the Schwarzschild geometry, tangential
    at~$\phi=0$.  Note the differences between these curves and the taut
    strings shown elsewhere: unlike taut strings, null geodesics may
    cross the event horizon inwards, and are never tangential to the
    event horizon}
  \label{light_closest_approach}
\end{figurehere}

@@

The concept of ``straight line'' is problematic in general relativity.
Although photon paths provide a definition of straightness, their
world lines are sensitive to the time component of the metric.  Light
taut strings respond only to the spatial components of a static
metric, and comparing their configuration to null geodesics is a
helpful way to think about straightness.  Using numerical methods to
compute string configuration is suitable for an undergraduate homework
problem.
